---
title: "The effect of digoxin on mortality and morbidity in patients with heart failure: a case of reproducible research"
author: Giacomo Bianchi
date: '2017-12-30'
type: "post"
output:
    html_document:
        theme: "cosmo"
        highlight: "default"
        toc: true
        toc_float: true
bibliography: biblio/digoxin.bib
csl: biblio/circulation.csl
slug: Digoxin NEJM
categories:
  - R
  - Data Analysis
  - Cardiology
tags:
  - Data Analysis
  - Tidy Data
  - Data Munging
  - R Notebook
---

#Philosophy
>Statistics: the only science that enables different experts using the same figures to draw different concluesion. **Evan Esar**

>The only statistics you can trust are those you falsified yourself. **Winston Churchill**


#Introduction and aim of the analysis
The aim of this analysis is to reproduce the landmark paper published by **The Digitalis Investigation Group** in 1997 in the *New England Journal of Medicine*[@digNEJM].
The original dataset has been released for teaching purpose by the [National Heart Lung and Blood Institute](https://www.nhlbi.nih.gov/).

#Perform an Exploratory Data Analysis 

```{r Load libraries for data analysis, message=FALSE, warning=FALSE, echo=FALSE}
library(tidyverse)
library(lubridate)
```

##Read the dataset 
The work of the data scientist or researcher begins with the exploration of the actual file that contains the data, its storage format and the way to read it.
The file is in my directory `/data` and i can access it using the command line interface:

```{bash CLI access to file}
cd data
cat dig.csv | head -n 2
```

The data are stored as "comma-separated-values" where the actual separator is, in fact, the "comma". It's a "tidy" dataset, where each row represent an patient and each column a variable.
Now it can be read and stored using the appropriate function from the "tidyverse" package `read_csv()`.

```{r Read the dataset into a dataframe}
dig <- read_csv(file = "data/dig.csv")
```

##Dataset structure
Now is time to explore the structure of the dataframe with the appropriate functions.

```{r Explore dig tibble}
glimpse(dig)
```


#Bibliography
